---
title: "Modeling Runners' Times in the Cherry Blossom Race - Case Study Unit 8"
output: html_notebook
---

####Cory Adams, Chris Boomhower, Alexandra Fisher, Alex Frye
####MSDS 7333, October 25, 2017

***NOTE: Answering Q.10: We have seen that the 1999 runners were typically older than the 2012 runners.
Compare the age distribution of the runners across all 14 years of the races. Use quantile-
quantile plots, boxplots, and density curves to make your comparisons. How do the
distributions change over the years? Was it a gradual change?***

##Abstract
In this case study, race results for the Cherry Blossom Ten Mile Run from 1999 to 2012 are used to study how people’s physical performance changes with age [1]. The focus of the study involves scraping the Web for free and publicly published data and then getting that data into the correct format. In order to investigate how age distributions change over the years, age distribution for all runners across 14 years is compared; however, what information is recorded and how the data is formatted changes each year so close attention is paid first to formatting the data. The R language is leveraged to appropriately format the data via statistical analysis and examining summary statistics and plots. Following data acquisition and successfully reading tables of race results into R, results are analyzed and compared using quantile-quantile plots, boxplots, and density curves. In this way, we are able to visualize the data for tens of thousands of observations to explore the performance-age relationship. [insert results / final outcome]

##Introduction
Race results for the Cherry Blossom Ten Mile Run are available to the public online at http://www.cherryblossom.org/. 

##Methods

##Results

```{r include=FALSE, cache=TRUE}
source('DataExtractAndClean.R', echo = FALSE)
```

####Load Libraries
```{r echo=FALSE, message=FALSE, warning = FALSE}
library(ggplot2)
library(plotly)
library(plyr)
library(grid)
library(gridExtra)
```


First, we remove runners under the age of 10 or with unrealistic run times (binning only meant to represent general distribution).
```{r}
cbMenSub = cbMen[cbMen$runTime > 30 & !is.na(cbMen$age) & cbMen$age >= 10, ]
ageCat = cut(cbMenSub$age, breaks = c(seq(10, 80, 10), 90))
cat("Binned age counts:\n")
table(ageCat)
```

```{r}
# Subset in order to color by year
cbMensPlot <- cbMenSub
cbMensPlot$year <- as.character(cbMensPlot$year)


age.d = ggplot(cbMensPlot, aes(age, color = year)) + geom_density()
age.d = ggplotly(age.d) #Convert to plotly version
age.d
```

```{r}
# calculate the normal theoretical quantiles per group
cbMensPlotQQ <- ddply(.data = cbMensPlot, .variables = .(year),
                      function(dsub){
                          q <- qqnorm(dsub$age, plot = FALSE)
                          dsub$xq <- q$x
                          dsub
                      })

age.qq = ggplot(data = cbMensPlotQQ, aes(x = xq, y = age, color = year)) +
                geom_point() +
                geom_smooth(method = "lm", se = FALSE) +
                xlab("Normal Theoretical Quantiles") +
                ylab("Normal Data Quantiles")
age.qq = ggplotly(age.qq)
age.qq
```

```{r fig.width=12, fig.asp=1.5}

plot.list <- list()
i = 1
#color.vect <- c('#D32F2F', '#C2185B', '#7B1FA2', '#512DA8', '#303F9F', '#1976D2', '#0288D1',
#                '#0097A7', '#00796B', '#388E3C', '#689F38', '#AFB42B', '#FBC02D', '#FFA000')

for(yr in unique(cbMensPlot$year)){
    cbMensPlotQQ <- ddply(.data = subset(cbMensPlot, cbMensPlot$year == yr), .variables = .(year),
                          function(dsub){
                              q <- qqnorm(dsub$age, plot = FALSE)
                              dsub$xq <- q$x
                              dsub
                          })
    
    age.qq = ggplot(data = cbMensPlotQQ, aes(x = xq, y = age, color = year)) +
                    geom_point() +
                    geom_smooth(method = "lm", se = FALSE) +
                    xlab("Normal Theoretical Quantiles") +
                    ylab("Normal Data Quantiles")
    
    #ggplot doesn't like to evaluate vector indexed contents so having to explicitly write color values line-by-line
    if(i==1) plot.list[[i]] <- age.qq + scale_color_manual(values='#D32F2F')
    if(i==2) plot.list[[i]] <- age.qq + scale_color_manual(values='#C2185B')
    if(i==3) plot.list[[i]] <- age.qq + scale_color_manual(values='#7B1FA2')
    if(i==4) plot.list[[i]] <- age.qq + scale_color_manual(values='#512DA8')
    if(i==5) plot.list[[i]] <- age.qq + scale_color_manual(values='#303F9F')
    if(i==6) plot.list[[i]] <- age.qq + scale_color_manual(values='#1976D2')
    if(i==7) plot.list[[i]] <- age.qq + scale_color_manual(values='#0288D1')
    if(i==8) plot.list[[i]] <- age.qq + scale_color_manual(values='#0097A7')
    if(i==9) plot.list[[i]] <- age.qq + scale_color_manual(values='#00796B')
    if(i==10) plot.list[[i]] <- age.qq + scale_color_manual(values='#388E3C')
    if(i==11) plot.list[[i]] <- age.qq + scale_color_manual(values='#689F38')
    if(i==12) plot.list[[i]] <- age.qq + scale_color_manual(values='#AFB42B')
    if(i==13) plot.list[[i]] <- age.qq + scale_color_manual(values='#FBC02D')
    if(i==14) plot.list[[i]] <- age.qq + scale_color_manual(values='#FFA000')
    
    #plot.list[[i]] <- age.qq + scale_color_manual(values=eval(as.character(color.vect[i])))
    i = i + 1
}

grid.arrange(plot.list[[1]], 
             plot.list[[2]],
             plot.list[[3]],
             plot.list[[4]],
             plot.list[[5]],
             plot.list[[6]],
             plot.list[[7]],
             plot.list[[8]],
             plot.list[[9]],
             plot.list[[10]],
             plot.list[[11]],
             plot.list[[12]],
             plot.list[[13]],
             plot.list[[14]], ncol=2)
```


##Discussion and Future Works

## References
[1] D. Lang and D. Nolan, Data Science in R: A Case Studies Approach to Computation Reasoning and Problem Solving. New York, New York: CRC Press. 
