---
title: "Modeling Runners' Times in the Cherry Blossom Race - Case Study Unit 8"
output: html_notebook
---

####Cory Adams, Chris Boomhower, Alexandra Fisher, Alex Frye
####MSDS 7333, October 25, 2017

***NOTE: Answering Q.10: We have seen that the 1999 runners were typically older than the 2012 runners.
Compare the age distribution of the runners across all 14 years of the races. Use quantile-
quantile plots, boxplots, and density curves to make your comparisons. How do the
distributions change over the years? Was it a gradual change?***

##Abstract

##Introduction

##Methods

##Results

```{r include=FALSE, cache=TRUE}
source('DataExtractAndClean.R', echo = FALSE)
```

####Load Libraries
```{r messages=FALSE}
library(ggplot2)
library(plotly)
library(manhattanly)
```


First, we remove runners under the age of 10 or with unrealistic run times.
```{r}
cbMenSub = cbMen[cbMen$runTime > 30 & !is.na(cbMen$age) & cbMen$age >= 10, ]
ageCat = cut(cbMenSub$age, breaks = c(seq(10, 80, 10), 90))
cat("Binned age counts:\n")
table(ageCat)
```

```{r}
# Subset in order to color by year
cbMenDens <- cbMenSub
cbMenDens$year <- as.character(cbMenDens$year)


age.d = ggplot(cbMenDens, aes(age, color = year)) + geom_density()
age.d = ggplotly(age.d) #Convert to plotly version
age.d
```




##Discussion and Future Works

## References
[1] D. Lang and D. Nolan, Data Science in R: A Case Studies Approach to Computation Reasoning and Problem Solving. New York, New York: CRC Press. 
